##Introduction  
Fine particulate matter (PM2.5) is an ambient air pollutant for which there is strong evidence that it is harmful to human health. In the United States, the Environmental Protection Agency (EPA) is tasked with setting national ambient air quality standards for fine PM and for tracking the emissions of this pollutant into the atmosphere. Approximatly every 3 years, the EPA releases its database on emissions of PM2.5. This database is known as the National Emissions Inventory (NEI). You can read more information about the NEI at the EPA National Emissions Inventory web site.  
  
For each year and for each type of PM source, the NEI records how many tons of PM2.5 were emitted from that source over the course of the entire year. The data that you will use for this assignment are for 1999, 2002, 2005, and 2008.

##Assignment  
The overall goal of this assignment is to explore the National Emissions Inventory database and see what it say about fine particulate matter pollution in the United states over the 10-year period 1999-2008. You may use any R package you want to support your analysis.


##Questions  
You must address the following questions and tasks in your exploratory analysis. For each question/task you will need to make a single plot. Unless specified, you can use any plotting system in R to make your plot.  
  
1. Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.  
2. Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips == "24510") from 1999 to 2008? Use the base plotting system to make a plot answering this question.  
3. Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999-2008 for Baltimore City? Which have seen increases in emissions from 1999-2008? Use the ggplot2 plotting system to make a plot answer this question.  
4. Across the United States, how have emissions from coal combustion-related sources changed from 1999-2008?  
5. How have emissions from motor vehicle sources changed from 1999-2008 in Baltimore City?  
6. Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (fips == "06037"). Which city has seen greater changes over time in motor vehicle emissions?  

---
  
####Load Libraries & Set Working Directory  
```{r LoadLibraries, echo=T, cache=T, eval=F}

#   Load Libraries
    library(ggplot2);
    library(utils);
    library(grid);
    library(plyr);

#   Set Working Directory
    setwd("~/Data Science/04 Exploratory Data Analysis/")

```
  
  
####Data Processing  
```{r DataProcessing1, echo=T, cache=T, eval=F}    

#   Downdload & Uncompress Data
    zip <- "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip";
    download.file(zip, destfile="~/Data Science/04 Exploratory Data Analysis/exdata_data_NEI_data.zip");
    unzip("~/Data Science/04 Exploratory Data Analysis/exdata_data_NEI_data.zip");

```

```{r DataProcessing2, echo=T, cache=T, eval=T} 

#   Import Data
    PM25 <- readRDS("~/Data Science/04 Exploratory Data Analysis/summarySCC_PM25.rds");
    SCC <- readRDS("~/Data Science/04 Exploratory Data Analysis/Source_Classification_Code.rds");

#   Exploratory Analysis
    length(PM25$Emissions);
    length(PM25$year);

```
  
  
####Question #1
Have total emissions from PM2.5 decreased in the United States from 1999 to 2008?  Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008. 
  
```{r Q1, echo=T, cache=T, eval=T}

#   Data Processing
    Total.PM25 <- tapply(PM25$Emissions, PM25$year, sum);

    plot(names(Total.PM25), Total.PM25, type="l", xlab="Year", ylab="Total Emissions",
         main="PM2.5 Emissions", col="blue", lwd=2, xlim=c(1998, 2008));

    dev.copy(png, file="plot1.png", width=480, height=480);
    dev.off();

```
  
**Answer:** Yes, total emissions from PM2.5 have decreased in the United States from 1999 to 2008.  ( See "Total PM2.5 Emissions" plot above. )
  
  
####Question #2
Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips == "24510") from 1999 to 2008?  Use the base plotting system to make a plot answering this question.  
  
```{r Q2, echo=T, cache=T, eval=T}

#   Data Processing
    baltimore <- subset (PM25, fips == "24510");
    total.PM25yr <- tapply(baltimore$Emissions, baltimore$year, sum);
    
#   Line Plot - "plot2.png"

    plot(names(total.PM25yr), total.PM25yr, type="l", xlab="Year", ylab="Total Emissions",
         main="PM2.5 Emissions", col="blue", lwd=2, xlim=c(1998, 2008));

    dev.copy(png, file="plot2.png", width=480, height=480);
    dev.off();

```

**Answer:** Yes, total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips == "24510") from 1999 to 2008.
  
  
####Question #3
Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999-2008 for Baltimore City? Which have seen increases in emissions from 1999-2008? Use the ggplot2 plotting system to make a plot answer this question. 
  
```{r Q3, echo=T, cache=T, eval=T}
    
#   Data Processing
    baltimore <- subset(PM25, fips == "24510");
    typePM25.year <- ddply(baltimore, .(year, type), function(x) sum(x$Emissions));
        colnames(typePM25.year)[3] <- "Emissions";
    
#   Line Plot - "plot3.png"
    plot3 <- qplot(year, Emissions, data=typePM25.year, color=type, geom="line");
    plot3 <- plot3 + ggtitle("Baltimore: PM2.5 Emmissions by Type");
    plot3 <- plot3 + xlab("Year") + ylab("Total Emissions");
    plot3;

    dev.copy(png, file="plot3.png", width=480, height=480);
    dev.off();

```

**Answer:** 
    
  
####Question #4
Across the United States, how have emissions from coal combustion-related sources changed from 1999 to 2008?
  
```{r Q4, echo=T, cache=T, eval=T}
    
#   Data Processing
    coalcomb.scc <- subset(SCC, EI.Sector %in% c("Fuel Comb - Comm/Instutional - Coal",
        "Fuel Comb - Electric Generation - Coal", "Fuel Comb - Industrial Boilers, ICEs Coal"));
    
    ## Step 3: comparisons so that we didn't ommit anything weird
    coalcomb.scc1 <- subset(SCC, grepl("Comb", Short.Name) & grepl("Coal", Short.Name))
    
    nrow(coalcomb.scc) #evaluate: rows 0
    ## [1] 35
    nrow(coalcomb.scc1) #evaluate: rows 91
    ## [1] 91
    ## Step 4: set the differences 
    dif1 <- setdiff(coalcomb.scc$SCC, coalcomb.scc1$SCC)
    dif2 <- setdiff(coalcomb.scc1$SCC, coalcomb.scc$SCC)
    
    length(dif1)#0
    ## [1] 6
    length(dif2)#91
    ## [1] 62
    ##Based on other coursera courses and previous history...
    ###it's time to look at the union of these sets
    coalcomb.codes <- union(coalcomb.scc$SCC, coalcomb.scc1$SCC)
    length(coalcomb.codes) #91
    ## [1] 97
    ## Step 5: subset again for what we want
    coal.comb <- subset(PM25, SCC %in% coalcomb.codes)
    
    ##Step 6: get the PM25 values as well
    coalcomb.pm25year <- ddply(coal.comb, .(year, type), function(x) sum(x$Emissions))
    colnames(coalcomb.pm25year)[3] <- "Emissions"

#   Line Plot - "plot4.png"
    plot4 <- qplot(year, Emissions, data=coalcomb.pm25year, color=type, geom="line");
    plot4 <- plot4 + stat_summary(fun.y="sum", fun.ymin="sum", fun.ymax="sum", color="purple", 
        aes(shape="total"), geom="line");
    plot4 <- plot4 + geom_line(aes(size="total", shape = NA));
    plot4 <- plot4 + ggtitle(expression("Coal Combustion" ~ PM[2.5] ~ "Emissions by Source Type and Year")); 
    plot4 <- plot4 + xlab("Year") + ylab(expression("Total" ~ PM[2.5] ~ "Emissions (tons)"));
    plot4;

    dev.copy(png, file="plot3.png", width=480, height=480);
    dev.off();

```
  
**Answer:** 
  
  
####Question #5
How have emissions from motor vehicle sources changed from 1999 - 2008 in Baltimore City?
  
```{r Q5, echo=T, cache=T, eval=T}
    
#   Data Processing
    mv.sourced <- unique(grep("Vehicles", SCC$EI.Sector, ignore.case = TRUE, value = TRUE))
    mv.sourcec <- SCC[SCC$EI.Sector %in% mv.sourced, ]["SCC"]
    
    #   Subset Vehicle Emissions Data
        emMV.ba <- NEI[NEI$SCC %in% mv.sourcec$SCC & NEI$fips == "24510",]
    
    #   Subset Vehicle Emissions Data for Baltimore
        balmv.pm25yr <- ddply(emMV.ba, .(year), function(x) sum(x$Emissions))
        colnames(balmv.pm25yr)[2] <- "Emissions"
    
#   Line Plot - "plot5.png"
    plot5 <- qplot(year, Emissions, data=balmv.pm25yr, geom="line") ;
    plot5 <- plot5 + ggtitle("Baltimore City PM2.5 Motor Vehicle Emissions by Year"); 
    plot5 <- plot5 + xlab("Year") + ylab("Total PM2.5 Emissions");
    plot5;

```
  
**Answer:** 
  
  
####Question #6
Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California, California (fips == "06037"). Which city has seen greater changes over time in motor vehicle emissions?
  
```{r Q6, echo=T, cache=T, eval=T}

#   Data Processing
    mv.sourced <- unique(grep("Vehicles", SCC$EI.Sector, ignore.case=T, value=T));
    mv.sourcec <- SCC[SCC$EI.Sector %in% mv.sourced, ]["SCC"];

    #   Subset Baltimore & Los Angeles County Data
        emMV.ba <- PM25[PM25$SCC %in% mv.sourcec$SCC & PM25$fips == "24510", ];
        emMV.LA <- PM25[PM25$SCC %in% mv.sourcec$SCC & PM25$fips == "06037", ];
        emMV.comb <- rbind(emMV.ba, emMV.LA);
    
    tmveYR.county <- aggregate (Emissions ~ fips * year, data=emMV.comb, FUN=sum) ;  # Vehicle Emissions
    tmveYR.county$county <- ifelse(tmveYR.county$fips=="06037", "Los Angeles", "Baltimore");  # Vehicle Emissions


#   Line Plot - "plot6.png"
    plot6 <- qplot(year, Emissions, data=tmveYR.county, geom="line", color=county);
    plot6 <- plot6 + ggtitle("Vehicle Emission Levels");
    plot6 <- plot6 + xlab("Year") + ylab("Levels of PM2.5 Emissions");
    plot6;

```
  
**Answer:** 
